---
title: "Airline Passenger Satisfaction – Additional Analysis"
author: "Rose and Anqi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{r load-data-libs}
# ── Load libraries ──────────────────────────────────────────────────────────
library(tidyverse)
library(ggplot2)
library(broom)
library(pROC)
library(car)        # vif()
library(ResourceSelection)  # hoslem.test()
library(corrplot)   # correlation heatmap
library(gridExtra)  # arrange multiple plots
library(scales)     # percent labels

# ── Load data ────────────────────────────────────────────────────────────────
# Update this path to match yours
airline <- read.csv("C:/Users/rosev/OneDrive/Documents/cleaned_airline_dataset.csv")
airline$satisfaction  <- trimws(airline$satisfaction)
airline$arrival_delay <- as.numeric(airline$arrival_delay)
airline$departure_delay <- as.numeric(airline$departure_delay)
```

---

# PART 1 — ENHANCED EXPLORATORY DATA ANALYSIS (EDA)

## 1.1  Dataset Summary

```{r eda-summary}
# Quick overview of all variables
glimpse(airline)
summary(airline)
```

## 1.2  Missing Values Check

```{r missing-values}
# Count NAs per column – important to report in any report
missing_counts <- colSums(is.na(airline))
missing_df <- data.frame(
  Variable    = names(missing_counts),
  Missing     = missing_counts,
  Pct_Missing = round(100 * missing_counts / nrow(airline), 2)
) %>% filter(Missing > 0)

if (nrow(missing_df) == 0) {
  cat("No missing values found in the dataset.\n")
} else {
  print(missing_df)
}
```

## 1.3  Outcome Distribution

```{r outcome-distribution}
# Proportion table + bar chart
prop.table(table(airline$satisfaction))

ggplot(airline, aes(x = satisfaction, fill = satisfaction)) +
  geom_bar(width = 0.5) +
  geom_text(stat = "count",
            aes(label = paste0(..count.., "\n(", round(..count.. / nrow(airline) * 100, 1), "%)")),
            vjust = -0.3, size = 4.5) +
  scale_fill_manual(values = c("satisfied" = "#2C7BB6", "neutral or dissatisfied" = "#D7191C")) +
  labs(title = "Distribution of Passenger Satisfaction",
       subtitle = "Slightly imbalanced: ~45% satisfied, ~55% neutral/dissatisfied",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## 1.4  Demographics & Passenger Profile

```{r demographics}
# Age distribution by satisfaction
p_age <- ggplot(airline, aes(x = age, fill = satisfaction)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("satisfied" = "#2C7BB6", "neutral or dissatisfied" = "#D7191C")) +
  labs(title = "Age Distribution by Satisfaction",
       x = "Age (years)", y = "Count", fill = "Satisfaction") +
  theme_minimal(base_size = 12)

# Customer type
p_ctype <- airline %>%
  count(customer_type, satisfaction) %>%
  group_by(customer_type) %>%
  mutate(pct = n / sum(n)) %>%
  filter(satisfaction == "satisfied") %>%
  ggplot(aes(x = customer_type, y = pct, fill = customer_type)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "% Satisfied by Customer Type",
       x = NULL, y = "% Satisfied") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Class
p_class <- airline %>%
  count(class, satisfaction) %>%
  group_by(class) %>%
  mutate(pct = n / sum(n)) %>%
  filter(satisfaction == "satisfied") %>%
  ggplot(aes(x = reorder(class, -pct), y = pct, fill = class)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "% Satisfied by Travel Class",
       x = NULL, y = "% Satisfied") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Travel type
p_type <- airline %>%
  count(type_of_travel, satisfaction) %>%
  group_by(type_of_travel) %>%
  mutate(pct = n / sum(n)) %>%
  filter(satisfaction == "satisfied") %>%
  ggplot(aes(x = type_of_travel, y = pct, fill = type_of_travel)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "% Satisfied by Travel Type",
       x = NULL, y = "% Satisfied") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

grid.arrange(p_age, p_ctype, p_class, p_type, ncol = 2)
```

## 1.5  Service Ratings: Distribution Heatmap

```{r service-heatmap}
# Average service ratings by satisfaction group – a clean heatmap
service_vars <- c("seat_comfort", "inflight_wifi_service", "food_and_drink",
                  "inflight_entertainment", "cleanliness", "leg_room_service",
                  "inflight_service", "online_boarding", "baggage_handling",
                  "checkin_service", "onboard_service")

service_means <- airline %>%
  select(satisfaction, all_of(service_vars)) %>%
  pivot_longer(-satisfaction, names_to = "service", values_to = "rating") %>%
  group_by(satisfaction, service) %>%
  summarise(mean_rating = round(mean(rating, na.rm = TRUE), 2), .groups = "drop")

ggplot(service_means, aes(x = satisfaction, y = service, fill = mean_rating)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = mean_rating), size = 4, fontface = "bold") +
  scale_fill_gradient2(low = "#D7191C", mid = "#FFFFBF", high = "#1A9641",
                       midpoint = 3, limits = c(1, 5),
                       name = "Avg Rating") +
  labs(title = "Average Service Ratings by Satisfaction Group",
       subtitle = "Green = higher rating | Red = lower rating",
       x = "Satisfaction", y = "Service Variable") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## 1.6  Service Ratings: Grouped Boxplots (All Services at Once)

```{r service-boxplots}
airline %>%
  select(satisfaction, all_of(service_vars)) %>%
  pivot_longer(-satisfaction, names_to = "service", values_to = "rating") %>%
  ggplot(aes(x = satisfaction, y = rating, fill = satisfaction)) +
  geom_boxplot(outlier.alpha = 0.3, outlier.size = 0.8) +
  facet_wrap(~ service, ncol = 4) +
  scale_fill_manual(values = c("satisfied" = "#2C7BB6", "neutral or dissatisfied" = "#D7191C")) +
  labs(title = "Service Rating Distributions by Satisfaction Group",
       x = NULL, y = "Rating (0–5)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        strip.text = element_text(size = 8))
```

## 1.7  Correlation Heatmap Among Service Variables

```{r correlation-heatmap}
# Check for multicollinearity visually BEFORE model
cor_matrix <- airline %>%
  select(all_of(service_vars)) %>%
  cor(use = "complete.obs")

corrplot(cor_matrix,
         method = "color",
         type   = "upper",
         tl.col = "black",
         tl.cex = 0.8,
         addCoef.col = "black",
         number.cex = 0.65,
         col = colorRampPalette(c("#D7191C", "white", "#2C7BB6"))(200),
         title = "Correlation Matrix: Service Quality Variables",
         mar   = c(0, 0, 2, 0))
```

## 1.8  Satisfaction Rate by Service Rating Level (All Services)

```{r satisfaction-by-rating}
# For each service, compute % satisfied at each rating level (0–5)
airline_clean <- airline %>%
  mutate(sat_binary = ifelse(satisfaction == "satisfied", 1, 0))

sat_by_rating <- airline_clean %>%
  select(sat_binary, all_of(service_vars)) %>%
  pivot_longer(-sat_binary, names_to = "service", values_to = "rating") %>%
  group_by(service, rating) %>%
  summarise(pct_satisfied = mean(sat_binary, na.rm = TRUE),
            n = n(), .groups = "drop")

ggplot(sat_by_rating, aes(x = rating, y = pct_satisfied, color = service, group = service)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = n), alpha = 0.7) +
  facet_wrap(~ service, ncol = 4) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_x_continuous(breaks = 0:5) +
  scale_size_continuous(range = c(1, 4), guide = "none") +
  labs(title = "Proportion Satisfied by Service Rating",
       subtitle = "Dot size = number of passengers at that rating",
       x = "Service Rating (0–5)", y = "% Satisfied") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        strip.text = element_text(size = 8))
```

## 1.9  Delay Distribution and Outlier Inspection

```{r delay-distribution}
p_dep <- ggplot(airline, aes(x = departure_delay)) +
  geom_histogram(bins = 60, fill = "#2C7BB6", alpha = 0.8) +
  scale_x_continuous(limits = c(0, 300)) +
  labs(title = "Departure Delay Distribution (0–300 min)",
       x = "Minutes", y = "Count") +
  theme_minimal(base_size = 12)

p_dep_log <- ggplot(airline %>% filter(departure_delay > 0),
                    aes(x = log1p(departure_delay))) +
  geom_histogram(bins = 40, fill = "#1A9641", alpha = 0.8) +
  labs(title = "Departure Delay – log(delay + 1) for Delayed Flights",
       x = "log(delay + 1)", y = "Count") +
  theme_minimal(base_size = 12)

grid.arrange(p_dep, p_dep_log, ncol = 2)

# Summary stats
cat("Departure delay summary:\n")
summary(airline$departure_delay)
cat("\nArrival delay summary:\n")
summary(airline$arrival_delay)
```

---

# PART 2 — LOGISTIC REGRESSION DIAGNOSTICS

> **This is the section the professor flagged.** Standard residuals / Q-Q plots
> from linear regression are *not* appropriate for logistic regression.  
> The correct diagnostics for GLM/logistic regression are:  
> (1) Binned residuals plot, (2) Pearson / deviance residuals, (3) Influence  
> measures (Cook's D), (4) Hosmer-Lemeshow goodness-of-fit test, and  
> (5) Linearity-in-log-odds check.

## 2.1  Rebuild Models on Full Data

```{r rebuild-models}
airline_clean <- airline %>%
  mutate(sat_binary = as.numeric(satisfaction == "satisfied")) %>%
  drop_na(all_of(service_vars), departure_delay, arrival_delay)

# Model 1: Service quality
model1 <- glm(sat_binary ~
                seat_comfort + inflight_wifi_service + food_and_drink +
                inflight_entertainment + cleanliness + leg_room_service +
                inflight_service + online_boarding + baggage_handling +
                checkin_service + onboard_service,
              data = airline_clean, family = binomial)

# Model 2: Delays (continuous)
model2 <- glm(sat_binary ~ departure_delay + arrival_delay,
              data = airline_clean, family = binomial)

cat("Model 1 – AIC:", AIC(model1), "\n")
cat("Model 2 – AIC:", AIC(model2), "\n")
```

## 2.2  Binned Residual Plot (Model 1) — PRIMARY DIAGNOSTIC

```{r binned-residuals-model1}
# Binned residuals: group predicted probabilities into bins, 
# plot average residual in each bin. Should lie within ±2 SE bands.
# Reference: Gelman & Hill (2007)

binned_residuals <- function(model, n_bins = 40, title = "Binned Residual Plot") {
  fitted  <- fitted(model)
  resid_p <- residuals(model, type = "pearson")
  
  df <- data.frame(fitted = fitted, resid = resid_p) %>%
    arrange(fitted) %>%
    mutate(bin = cut(fitted, breaks = n_bins, labels = FALSE))
  
  bin_summary <- df %>%
    group_by(bin) %>%
    summarise(
      mean_fitted = mean(fitted),
      mean_resid  = mean(resid),
      se          = sd(resid) / sqrt(n()),
      .groups = "drop"
    )
  
  se_bound <- 2 / sqrt(nrow(df) / n_bins)
  
  ggplot(bin_summary, aes(x = mean_fitted, y = mean_resid)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_ribbon(aes(ymin = -se_bound, ymax = se_bound),
                fill = "#AED6F1", alpha = 0.4) +
    geom_point(color = "#2C7BB6", size = 2) +
    geom_line(color = "#2C7BB6", alpha = 0.5) +
    labs(title = title,
         subtitle = "Points should fall within shaded ±2 SE band for good fit",
         x = "Fitted Probability", y = "Average Pearson Residual") +
    theme_minimal(base_size = 12)
}

binned_residuals(model1, title = "Binned Residuals – Model 1 (Service Quality)")
```

## 2.3  Binned Residual Plot (Model 2)

```{r binned-residuals-model2}
binned_residuals(model2, title = "Binned Residuals – Model 2 (Delays)")
```

## 2.4  Pearson & Deviance Residuals vs Fitted (Model 1)

```{r residuals-vs-fitted}
aug1 <- augment(model1, type.predict = "response", type.residuals = "pearson")

p_pearson <- ggplot(aug1, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, color = "#2C7BB6", size = 0.8) +
  geom_smooth(se = FALSE, color = "red", method = "loess") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Pearson Residuals vs Fitted – Model 1",
       x = "Fitted Probability", y = "Pearson Residual") +
  theme_minimal(base_size = 12)

aug1_dev <- augment(model1, type.predict = "response", type.residuals = "deviance")

p_deviance <- ggplot(aug1_dev, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, color = "#1A9641", size = 0.8) +
  geom_smooth(se = FALSE, color = "red", method = "loess") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Deviance Residuals vs Fitted – Model 1",
       x = "Fitted Probability", y = "Deviance Residual") +
  theme_minimal(base_size = 12)

grid.arrange(p_pearson, p_deviance, ncol = 2)
```

## 2.5  Cook's Distance — Influential Observations

```{r cooks-distance}
cooks_d <- cooks.distance(model1)
obs_idx <- 1:length(cooks_d)
threshold <- 4 / nrow(airline_clean)

cook_df <- data.frame(obs = obs_idx, cooks_d = cooks_d)

ggplot(cook_df, aes(x = obs, y = cooks_d)) +
  geom_col(aes(fill = cooks_d > threshold), width = 1, show.legend = FALSE) +
  geom_hline(yintercept = threshold, linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("FALSE" = "#AED6F1", "TRUE" = "#D7191C")) +
  annotate("text", x = max(obs_idx) * 0.85, y = threshold * 1.2,
           label = paste0("Threshold = 4/n = ", round(threshold, 4)),
           color = "red", size = 3.5) +
  labs(title = "Cook's Distance – Model 1",
       subtitle = "Red bars = potentially influential observations",
       x = "Observation Index", y = "Cook's Distance") +
  theme_minimal(base_size = 12)

cat("Number of influential points (Cook's D > 4/n):",
    sum(cooks_d > threshold), "\n")
```

## 2.6  Hosmer-Lemeshow Goodness-of-Fit Test

```{r hosmer-lemeshow}
# H0: model fits well; p > 0.05 → no evidence of poor fit
hl_test <- hoslem.test(airline_clean$sat_binary, fitted(model1), g = 10)
print(hl_test)

cat("\nInterpretation:\n")
if (hl_test$p.value > 0.05) {
  cat("p =", round(hl_test$p.value, 4),
      "→ Fail to reject H0. No significant evidence of poor fit.\n")
} else {
  cat("p =", round(hl_test$p.value, 4),
      "→ Reject H0. Model may not fit well — consider adding interactions or nonlinear terms.\n")
}
```

## 2.7  Linearity in Log-Odds (Box-Tidwell Check)

```{r log-odds-linearity}
# For logistic regression, numeric predictors should have a linear 
# relationship with the LOG-ODDS (not the probability directly).
# We check this by plotting the empirical log-odds in bins.

check_log_odds_linearity <- function(var, data, sat_col = "sat_binary",
                                     n_bins = 10, label = NULL) {
  df <- data %>%
    select(y = !!sym(sat_col), x = !!sym(var)) %>%
    drop_na() %>%
    filter(x > 0) %>%               # remove 0-rating if desired
    mutate(bin = ntile(x, n_bins)) %>%
    group_by(bin) %>%
    summarise(
      mean_x  = mean(x),
      p_sat   = mean(y),
      .groups = "drop"
    ) %>%
    mutate(log_odds = log(p_sat / (1 - p_sat)))
  
  ggplot(df, aes(x = mean_x, y = log_odds)) +
    geom_point(size = 2.5, color = "#2C7BB6") +
    geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
    labs(title = paste("Log-Odds Linearity Check:", ifelse(is.null(label), var, label)),
         x = var, y = "Empirical Log-Odds") +
    theme_minimal(base_size = 11)
}

# Check the two strongest predictors + wifi
p1 <- check_log_odds_linearity("online_boarding",        airline_clean)
p2 <- check_log_odds_linearity("leg_room_service",       airline_clean)
p3 <- check_log_odds_linearity("inflight_wifi_service",  airline_clean)
p4 <- check_log_odds_linearity("inflight_entertainment", airline_clean)

grid.arrange(p1, p2, p3, p4, ncol = 2,
             top = "Linearity in Log-Odds – Key Predictors\n(Points should follow the red line)")
```

---

# PART 3 — ENHANCED MODEL VISUALIZATIONS

## 3.1  Odds Ratio Forest Plot (Improved)

```{r or-forest-plot}
tidy_m1 <- tidy(model1, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = p.value < 0.05,
    term_label = case_when(
  term == "online_boarding"        ~ "Online Boarding",
  term == "leg_room_service"       ~ "Leg Room Service",
  term == "onboard_service"        ~ "Onboard Service",
  term == "inflight_entertainment" ~ "Inflight Entertainment",
  term == "cleanliness"            ~ "Cleanliness",
  term == "checkin_service"        ~ "Check-in Service",
  term == "baggage_handling"       ~ "Baggage Handling",
  term == "inflight_wifi_service"  ~ "Inflight Wi-Fi",
  term == "seat_comfort"           ~ "Seat Comfort",
  term == "inflight_service"       ~ "Inflight Service",
  term == "food_and_drink"         ~ "Food & Drink",
  TRUE                             ~ term
)
  )

ggplot(tidy_m1,
       aes(x = reorder(term_label, estimate),
           y = estimate,
           ymin = conf.low,
           ymax = conf.high,
           color = significant)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  geom_errorbar(width = 0.3, linewidth = 0.8) +
  geom_point(size = 3) +
  coord_flip() +
  scale_color_manual(values = c("TRUE" = "#D7191C", "FALSE" = "#AAAAAA"),
                     labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05")) +
  labs(
    title    = "Odds Ratios: Service Quality Predictors of Satisfaction",
    subtitle = "Model 1 — Logistic Regression (n = 1,000)\nOR > 1: associated with higher odds of satisfaction",
    x        = "Predictor",
    y        = "Odds Ratio (95% CI)",
    color    = "Significance"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## 3.2  Predicted Probability Plot: Online Boarding

```{r predicted-prob-online-boarding}
# Hold all other variables at their median, vary online_boarding
new_data_ob <- expand.grid(
  online_boarding        = 0:5,
  seat_comfort           = median(airline_clean$seat_comfort, na.rm = TRUE),
  inflight_wifi_service  = median(airline_clean$inflight_wifi_service, na.rm = TRUE),
  food_and_drink         = median(airline_clean$food_and_drink, na.rm = TRUE),
  inflight_entertainment = median(airline_clean$inflight_entertainment, na.rm = TRUE),
  cleanliness            = median(airline_clean$cleanliness, na.rm = TRUE),
  leg_room_service       = median(airline_clean$leg_room_service, na.rm = TRUE),
  inflight_service       = median(airline_clean$inflight_service, na.rm = TRUE),
  baggage_handling       = median(airline_clean$baggage_handling, na.rm = TRUE),
  checkin_service        = median(airline_clean$checkin_service, na.rm = TRUE),
  onboard_service        = median(airline_clean$onboard_service, na.rm = TRUE)
)

pred_ob <- predict(model1, newdata = new_data_ob, type = "response", se.fit = TRUE)
new_data_ob$pred   <- pred_ob$fit
new_data_ob$lwr    <- pred_ob$fit - 1.96 * pred_ob$se.fit
new_data_ob$upr    <- pred_ob$fit + 1.96 * pred_ob$se.fit

ggplot(new_data_ob, aes(x = online_boarding, y = pred)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, fill = "#2C7BB6") +
  geom_line(color = "#2C7BB6", linewidth = 1.2) +
  geom_point(color = "#2C7BB6", size = 3) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_x_continuous(breaks = 0:5) +
  labs(
    title    = "Predicted Probability of Satisfaction vs. Online Boarding Rating",
    subtitle = "All other predictors held at their median | Shaded area = 95% CI",
    x        = "Online Boarding Rating (0–5)",
    y        = "Predicted Probability of Satisfaction"
  ) +
  theme_minimal(base_size = 13)
```

## 3.3  Predicted Probability: Leg Room Service

```{r predicted-prob-leg-room}
new_data_lr <- expand.grid(
  leg_room_service       = 0:5,
  seat_comfort           = median(airline_clean$seat_comfort, na.rm = TRUE),
  inflight_wifi_service  = median(airline_clean$inflight_wifi_service, na.rm = TRUE),
  food_and_drink         = median(airline_clean$food_and_drink, na.rm = TRUE),
  inflight_entertainment = median(airline_clean$inflight_entertainment, na.rm = TRUE),
  cleanliness            = median(airline_clean$cleanliness, na.rm = TRUE),
  online_boarding        = median(airline_clean$online_boarding, na.rm = TRUE),
  inflight_service       = median(airline_clean$inflight_service, na.rm = TRUE),
  baggage_handling       = median(airline_clean$baggage_handling, na.rm = TRUE),
  checkin_service        = median(airline_clean$checkin_service, na.rm = TRUE),
  onboard_service        = median(airline_clean$onboard_service, na.rm = TRUE)
)

pred_lr <- predict(model1, newdata = new_data_lr, type = "response", se.fit = TRUE)
new_data_lr$pred <- pred_lr$fit
new_data_lr$lwr  <- pred_lr$fit - 1.96 * pred_lr$se.fit
new_data_lr$upr  <- pred_lr$fit + 1.96 * pred_lr$se.fit

ggplot(new_data_lr, aes(x = leg_room_service, y = pred)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, fill = "#D7191C") +
  geom_line(color = "#D7191C", linewidth = 1.2) +
  geom_point(color = "#D7191C", size = 3) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_x_continuous(breaks = 0:5) +
  labs(
    title    = "Predicted Probability of Satisfaction vs. Leg Room Service Rating",
    subtitle = "All other predictors held at their median | Shaded area = 95% CI",
    x        = "Leg Room Service Rating (0–5)",
    y        = "Predicted Probability of Satisfaction"
  ) +
  theme_minimal(base_size = 13)
```

## 3.4  ROC Curve – Publication Quality

```{r roc-curve}
set.seed(123)
n <- nrow(airline_clean)
train_idx <- sample(1:n, 0.8 * n)
train_df  <- airline_clean[train_idx, ]
test_df   <- airline_clean[-train_idx, ]

model_train <- glm(sat_binary ~
                     online_boarding + leg_room_service + onboard_service +
                     inflight_entertainment + cleanliness + checkin_service +
                     baggage_handling + inflight_wifi_service + seat_comfort +
                     inflight_service + food_and_drink,
                   data = train_df, family = binomial)

test_df$prob <- predict(model_train, newdata = test_df, type = "response")

roc_obj <- roc(test_df$sat_binary, test_df$prob, quiet = TRUE)
auc_val <- auc(roc_obj)

roc_df <- data.frame(
  sensitivity = roc_obj$sensitivities,
  specificity = roc_obj$specificities
)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#2C7BB6", linewidth = 1.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  geom_area(fill = "#AED6F1", alpha = 0.3) +
  annotate("text", x = 0.65, y = 0.25,
           label = paste0("AUC = ", round(auc_val, 3)),
           size = 5.5, fontface = "bold", color = "#2C7BB6") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "ROC Curve – Model 1 (Service Quality)",
    subtitle = "Test set (20% holdout) | Dashed line = random classifier",
    x        = "1 – Specificity (False Positive Rate)",
    y        = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal(base_size = 13)
```

## 3.5  Confusion Matrix Heatmap

```{r confusion-matrix}
test_df$pred_class <- ifelse(test_df$prob > 0.5, 1, 0)
cm <- table(Predicted = test_df$pred_class, Actual = test_df$sat_binary)
cm_df <- as.data.frame(cm) %>%
  mutate(
    label    = c("True Negative\n(TN)", "False Positive\n(FP)",
                 "False Negative\n(FN)", "True Positive\n(TP)"),
    fill_col = c("correct", "wrong", "wrong", "correct")
  )

ggplot(cm_df, aes(x = factor(Actual), y = factor(Predicted), fill = fill_col)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(label, "\n\nn = ", Freq)),
            size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("correct" = "#2C7BB6", "wrong" = "#D7191C")) +
  scale_x_discrete(labels = c("0" = "Neutral/\nDissatisfied", "1" = "Satisfied")) +
  scale_y_discrete(labels = c("0" = "Predicted:\nNeutral", "1" = "Predicted:\nSatisfied")) +
  labs(
    title    = "Confusion Matrix – Model 1 (Test Set)",
    subtitle = paste0("Accuracy = ",
                      round(mean(test_df$pred_class == test_df$sat_binary) * 100, 1), "%"),
    x        = "Actual",
    y        = "Predicted"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# PART 4 — MODEL 2 DELAY DIAGNOSTICS

## 4.1  Binned Residuals for Model 2

```{r model2-binned}
binned_residuals(model2, title = "Binned Residuals – Model 2 (Delays Only)")
```

## 4.2  Hosmer-Lemeshow for Model 2

```{r model2-hosmer}
hl2 <- hoslem.test(airline_clean$sat_binary, fitted(model2), g = 10)
print(hl2)
cat("\np-value =", round(hl2$p.value, 4), "\n")
```

## 4.3  Delay Effect Plot with Smoothed Probability

```{r delay-effect-plot}
airline_clean %>%
  filter(departure_delay <= 200) %>%
  ggplot(aes(x = departure_delay, y = sat_binary)) +
  geom_jitter(height = 0.04, alpha = 0.2, color = "steelblue", size = 0.7) +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              color = "red", se = TRUE, linewidth = 1.2) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "Departure Delay vs. Predicted Probability of Satisfaction",
    subtitle = "Red curve = logistic regression fit | Shaded = 95% CI | Delays ≤ 200 min shown",
    x        = "Departure Delay (minutes)",
    y        = "Probability of Satisfaction"
  ) +
  theme_minimal(base_size = 13)
```

---

# PART 5 — SUMMARY TABLE (Report-Ready)

```{r summary-table}
# Clean model summary table for the report
tidy_m1 %>%
  select(term_label, estimate, conf.low, conf.high, p.value, significant) %>%
  arrange(desc(estimate)) %>%
  mutate(
    OR       = round(estimate, 3),
    CI_lower = round(conf.low, 3),
    CI_upper = round(conf.high, 3),
    p_value  = ifelse(p.value < 0.001, "< 0.001", round(p.value, 3)),
    Sig      = ifelse(significant, "✓", "")
  ) %>%
  select(Predictor = term_label, OR, CI_lower, CI_upper, p_value, Sig) %>%
  knitr::kable(caption = "Model 1 – Odds Ratios, 95% CIs, and p-values")
```

```{r session-info, echo=FALSE}
cat("\n\nSession Info:\n")
sessionInfo()
```
